Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-09-27 arxiv/proof_examples.tex
wip
revise

\subsection{Proof of the Special Case}\label{app_iso_cov}
\todo{switch the $a_3$ and $a_4$}

@@

\begin{proof}[Proof of Proposition \ref{prop_dist_transition}]
	In the setting of Proposition \ref{prop_dist_transition}, we have $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} = \id$. Then solving equations \eqref{eq_a12extra} and \eqref{eq_a34extra} with $\lambda_i\equiv 1$, we get that
	\begin{align}
		 a_1 = \frac{\rho_1(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} ,\quad
		& a_2 = \frac{\rho_2(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} , \label{simplesovlea12}\\
		 a_3 = \frac{\rho_2}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}, \quad
		& a_4 = \frac{\rho_1}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}.\label{simplesovlea34}
	\end{align}
	Using Lemma \ref{lem_minv} and Lemma \ref{lem_cov_shift}, we can track the reduction of variance from $\hat{\beta}_2^{\MTL}$ to $\hat{\beta}_2^{\STL}$ as 
\be\label{Deltavar}
\begin{split}
\delta_{\vari}&:=\sigma^2  \bigtr{(X_2^{\top}X_2)^{-1} }  - (1+\OO(n^{-\e}))\cdot \sigma^2  \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} } =\Delta_{\vari}\cdot (1+\OO(n^{-\e})) 
\end{split}
\ee
with high probability, where 
	\begin{align*}
		\Delta_{\vari} &\define \sigma^2 \bigbrace{\frac{1}{\rho_2 - 1} - \frac{1}{\rho_1 + \rho_2}\cdot\frac{1}{a_1+ a_2} } =\sigma^2  \cdot \frac{\rho_1}{(\rho_2-1)(\rho_1 + \rho_2 -1)}.
	\end{align*}
	%where we use equation \eqref{eq_a12} and Lemma \ref{lem_hat_v}.
	Next for the model shift bias
	$$\delta_{\bias}:=(1+\OO(n^{-\e}))\cdot d^2 \tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right], $$
	we can get from Lemma \ref{prop_model_shift_tight} that
\be\label{Deltabeta} 
\al_-^2(\rho_1) - \oo(1)  \le \frac{\delta_{\bias}}{ \Delta_{\bias}} \le \al_+^2(\rho_1) +  \oo(1) , \ee
	where 
	$$\Delta_{\bias}:=pd^2 \cdot \frac{\rho_1^2}{(\rho_1+\rho_2)^2}  \frac{1 + a_3 + a_4}{(a_1 + a_2)^2}= pd^2 \cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}.$$	
%\begin{align*}
%		\Delta_{\bias} &\define \hat{v}^2 \bignorm{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1 (\beta_s - \hat{v}\beta_t)}^2 \\
%		&= d^2 \cdot \bignormFro{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}X_1^{\top}X_1}^2 + \bigo{p^{-1/2 + \varepsilon}d^2}
%	\end{align*}
%	Using Lemma \ref{prop_model_shift_tight}, we get an upper and lower bound on $\Delta_{\bias}$ as
%	\be\label{Deltabeta} \bigbrace{1 - \sqrt{\frac{1}{c_1}}}^4 \le \Delta_{\bias} / \bigbrace{p\cdot d^2 \cdot \frac{c_1^2}{(c_1+c_2)^2} \cdot \frac{1 + a_3 + a_4}{(a_1 + a_2)^2} + \bigo{p^{1/2+\e}d^2} } \le \bigbrace{1 + \sqrt{\frac{1}{c_1}}}^4. \ee	
%	Hence we obtain that
%	\begin{align*}
%		\frac{1 + a_3 + a_4}{(a_1 + a_2)^2}
%		= \frac{(c_1 + c_2)^3}{(c_1 + c_2 - 1)^3} + \bigo{p^{-1/2+\varepsilon}}.
%	\end{align*}
Note that by \eqref{simple1}, we have
 \be\label{var-beta}\te(\hat{\beta}_2^{\STL})-\te(\hat{\beta}_2^{\MTL}) =\delta_{\vari} - \delta_{\bias} .\ee
Then we can track its sign using \eqref{Deltavar} and \eqref{Deltabeta}.

>>>>

\subsection{The Random-Effects Model}\label{app_iso_cov}
We now give the proof of Proposition \ref{cor_MTL_loss}. As in the proof of Theorem \ref{thm_many_tasks}, we first derive a bias-variance decomposition of the prediction loss $L(\hat \eta_2^{\MTL})$. Then we can calculate the bias term and the variance term using Theorem \ref{thm_main_RMT}.

Under the two-task setting of Corollary \ref{cor_MTL_loss}, the optimization objective \eqref{eq_mtl} becomes
	\begin{align}
		f(A, B) =   \bignorm{X^{(1)} B A_1 - Y^{(1)}}^2+ \bignorm{X^{(2)} B A_2 - Y^{(2)}}^2, \label{eq_mtl_2task_cov}
	\end{align}
	where $B \in \real^{p}$ and $A_1, A_2\in \R$ because $r=1$ in the two-task case. % for $1 < r < t$ by our assumption.
	Without loss of generality, we assume that both $A_1$ and $A_2$ are nonzero, since otherwise we can add a small perturbation $\delta$ to them and later remove this perturbation by taking $\delta\to 0$ in the final result. Again using the local optimality condition ${\partial f}/{\partial B} = 0$, for any $A_1$ and $A_2$ we obtain the optimal $\hat{B}$ as follows
	\begin{align}
		\hat{B} \define  (A_1^2 (X^{(1)})^{\top}X^{(1)} + A_2^2 (X^{(2)})^{\top}X^{(2)})^{-1} (A_1 (X^{(1))})^{\top}Y^{(1)} + A_2 (X^{(2)})^{\top}Y^{(2)}). \label{eq_Bhat_2task} %\\
		%&= (B^\star A ^{\top}) (A A^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{j=1}^t \varepsilon_i A_i^{\top}} (A  A^{\top})^{-1}.
	\end{align}
 Now inserting $\hat B$ back into \eqref{eq_mtl_2task_cov}, we obtain an objective that only depends on $A:=A_1/A_2$ as follows 
 \begin{align}
		g(A) \define  \sum_{i=1}^2 . \label{eq_mtl_A12}
	\end{align}

\cor
%Our key insight is a bias-variance decomposition of the expected prediction loss of $\hat{\beta}_2^{\MTL} = \hat{B} W_2$.
%Using the local optimality condition of $B$, we obtain that
%\begin{align*}
%	 \hat{B}(W_1, W_2) &= (W_1^2 X_1^{\top}X_1 + W_2^2 X_2^{\top}X_2)^{-1} (W_1 X_1^{\top}Y_1 + W_2 X_2^{\top}Y_2)\\
%	&= \frac{1}{W_2} \left( \frac{W_1^2}{W_2^2}  X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1} \left(\frac{W_1}{W_2} X_1^{\top}Y_1 + X_2^{\top}Y_2\right) \\
%	&= \frac{1}{W_2}\left[\beta_2 + \left(\frac{W_1^2}{W_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}\bigbrace{X_1^{\top}X_1\left(\frac{W_1}{W_2}\beta_1 - \frac{W_1^2}{W_2^2} \beta_2\right) + \left(\frac{W_1}{W_2} X_1^{\top}\varepsilon_1 + X_2^{\top}\varepsilon_2\right)}\right].
%\end{align*}
Recall that $\hat{\beta}_i^{\MTL} = \hat{B} W_i$. Therefore, we have
\begin{align}
	\exarg{\epsilon_1, \epsilon_2}{L(\hat{\beta}_2^{\MTL}) \mid X_1, X_2}
	=&~ \frac{W_1^2}{W_2^2} \bignorm{\Sigma_2^{1/2}(\frac{W_1^2}{W_2^2} X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1 (\beta_1 - \frac{W_1}{W_2} \beta_2)}^2 \label{eq_bias_2task} \\
			&+~  \sigma^2\cdot \bigtr{\Sigma_2(\frac{W_1^2}{W_2^2} X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} }. \label{eq_var_2task}
\end{align}
Equation \eqref{eq_bias_2task} is the bias of $\hat{\beta}_t^{\MTL}$ and
equation \eqref{eq_var_2task} is the variance of $\hat{\beta}_t^{\MTL}$.
%minus the variance of $\hat{\beta}_t^{\STL}$, which is always negative.
Comparing the above to single-task learning, that is,
\begin{align}
	\exarg{\epsilon_2}{L(\hat{\beta}_2^{\STL}) \mid X_2} = \sigma^2 \cdot \bigtr{\Sigma_2 (X_2^{\top} X_2)^{-1}}, \label{eq_var_stl}
\end{align}
we observe that while the bias of $\hat{\beta}_2^{\MTL}$ is always larger than that of $\hat{\beta}_2^{\STL}$, which is zero, the variance of $\hat{\beta}_2^{\MTL}$ is always lower than that of $\hat{\beta}_2^{\STL}$.\footnote{To see why this is true, we apply the Woodbury matrix identity over equation \eqref{eq_var_2task} and use the fact that for the product of two PSD matrices, its trace is always nonnegative.}
In other words, training both tasks together helps predict the target task by reducing variance while incurring a bias.
Therefore, whether multi-task learning outperforms single-task learning is determined by the bias-variance decomposition!

% \todo{derive the test loss} Expectation was given in \eqref{eq_bias_2task} and \eqref{eq_var_2task}.
 \nc

Plugging $Y^{(1)}$ and $Y^{(2)}$ in \eqref{model_YvsX} into \eqref{eq_Bhat_2task}, we can rewrite that 
 \begin{align}
		\hat{B}= \frac{1}{A_2}\left[\beta_2 + \left(\frac{A_1^2}{A_2^2}(X^{(1)})^{\top}X^{(1)} +  (X^{(2)})^{\top}X^{(2)}\right)^{-1}\bigbrace{ (X^{(1)})^{\top}X_1\left(\frac{A_1}{A_2}\beta_1 - \frac{A_1^2}{A_2^2} \beta_2\right) + \left(\frac{A_1}{A_2} (X^{(1)})^{\top}\varepsilon_1 + (X^{(2)})^{\top}\varepsilon_2\right)}\right]. \label{eq_Bhat_2task2} %\\
		%&= (B^\star A ^{\top}) (A A^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{j=1}^t \varepsilon_i A_i^{\top}} (A  A^{\top})^{-1}.
	\end{align}
\todo{next derive the prediction loss} 
 \cob

 \begin{claim}
	In the setting described above, we have that
	\be\label{approxvalid}
		g(v)=  \left[\val(v) + (N_1\sigma^2_1+N_2\sigma^2_2)\right]\cdot \left( 1+\OO(p^{-(1-\e_0)/2+\e})\right).
	\ee
\end{claim}
\begin{proof}
	We use the fact that our random vectors have i.i.d. entries.
%Before doing that, we first need to fix the setting for the following discussions, because we want to keep track of the error rate carefully instead of obtaining an asymptotic result only.
	Recall that $Y_i = X_i\beta_i + \varepsilon_i$ and $\wt Y_i = \wt X_i\beta_i + \wt\varepsilon_i$, $i=1,2$, all satisfy Assumption \ref{assm_secA2}. Then we rewrite \eqref{eq_mtl_2tasktilde} as
$$	g( v) = \sum_{i=1}^2\left\| \wt X_i\wt\beta_i  - \wt \e_i\right\|^2 , \quad \wt\beta:=\hat B w_i-\beta_i.$$
Since $ \wt X_i\wt\beta$ and $ \wt \e_i$ are independent random vectors with i.i.d. centered entries, we can use the concentration result, Lemma \ref{largedeviation}, to get that for any constant $\e>0$,
\begin{align*}
\left|\left\| \wt X_i\wt\beta_i  - \wt \e_i\right\|^2 -  \exarg{\wt X_i,\wt{\e}_i} {\left\| \wt X_i\wt\beta_i  - \wt \e_i\right\|^2} \right| & =\left|\left\| \wt X_i\wt\beta_i  - \wt \e_i\right\|^2 - N_i (\wt\beta_i^\top \Sigma_i \wt\beta_i + \sigma_i^2) \right| \\
&\le N_i^{1/2+\e} (\wt\beta_i^\top \Sigma_i \wt\beta_i + \sigma_i^2),
\end{align*}
with high probability. Thus we obtain that
$$g(v)= \left[\sum_{i=1}^2 N_i\left\|\Sigma_i^{1/2}( \hat B w_i - \beta_i) \right\|^2 + (N_1\sigma^2_1+N_2\sigma^2_2)\right]\cdot \left( 1+\OO(p^{-(1-\e_0)/2+\e})\right),$$
where we also used $N_i\ge p^{-1+\e_0}$. Inserting \eqref{hatB} into the above expression and using
 again the concentration result, Lemma \ref{largedeviation}, we get that
$$ \sum_{i=1}^2 N_i\left\|\Sigma_i^{1/2}( \hat B w_1 - \beta_i) \right\|^2 = \val(v)\cdot \left( 1+\OO(p^{-1/2+\e})\right)$$
with high probability.
%-----old-------
%Suppose that the entries of $\e_1$ and $\e_2$ have variance $\sigma^2$.  Using a validation set that is sub-sampled from the original training dataset, we get a validation loss as follows
%\begin{align}
%		&\val(\hat{B}; w_1, w_2):= \exarg{\varepsilon_1,\e_2} \sum_{i=1}^2 \left\|\Sigma_i^{1/2}( \hat B w_1 - \beta_i) \right\|^2 \\
%	&=  n_1 \cdot \bignorm{\Sigma_1^{1/2}\left(\frac{w_1^2}{w_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}X_2^{\top}X_2\left (\beta_s - \frac{w_1}{w_2}\beta_t\right)}^2 \nonumber \\
%		&+ n_1 \sigma^2 \cdot \frac{w_1^2}{w_2^2} \bigtr{\left(\frac{w_1^2}{w_2^2}  X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}\Sigma_1} \nonumber \\
%		&+ n_2 \cdot \frac{w_1^2}{w_2^2}\bignorm{\Sigma_2^{1/2}\left(\frac{w_1^2}{w_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}X_1^{\top}X_1\left(\beta_s - \frac{w_1}{w_2}\beta_t\right)}^2 \nonumber \\
%		&+ n_2 \sigma^2 \cdot \bigtr{\left(\frac{w_1^2}{w_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}\Sigma_2}. \label{eq_val_mtl}
%\end{align}
%\nc
%------------------
Thus we conclude the proof.
\end{proof}

\nc

% \todo{switch the $a_3$ and $a_4$}

@@

\begin{proof}[Proof of Proposition \ref{prop_dist_transition}]
	In the setting of Proposition \ref{prop_dist_transition}, we have $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} = \id$. Then solving equations \eqref{eq_a12extra} and \eqref{eq_a34extra} with $\lambda_i\equiv 1$, we get that
	\begin{align}
		 a_1 = \frac{\rho_1(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} ,\quad
		& a_2 = \frac{\rho_2(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} , \label{simplesovlea12}\\
		a_3 = \frac{\rho_1}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}, \quad
		&  a_4 = \frac{\rho_2}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}.\label{simplesovlea34}
	\end{align}
	Using Lemma \ref{lem_minv} and Lemma \ref{lem_cov_shift}, we can track the reduction of variance from $\hat{\beta}_2^{\MTL}$ to $\hat{\beta}_2^{\STL}$ as 
\be\label{Deltavar}
\begin{split}
\delta_{\vari}&:=\sigma^2  \bigtr{(X_2^{\top}X_2)^{-1} }  - (1+\OO(n^{-\e}))\cdot \sigma^2  \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} } =\Delta_{\vari}\cdot (1+\OO(n^{-\e})) 
\end{split}
\ee
with high probability, where 
	\begin{align*}
		\Delta_{\vari} &\define \sigma^2 \bigbrace{\frac{1}{\rho_2 - 1} - \frac{1}{\rho_1 + \rho_2}\cdot\frac{1}{a_1+ a_2} } =\sigma^2  \cdot \frac{\rho_1}{(\rho_2-1)(\rho_1 + \rho_2 -1)}.
	\end{align*}
	%where we use equation \eqref{eq_a12} and Lemma \ref{lem_hat_v}.
	Next for the model shift bias
	$$\delta_{\bias}:=(1+\OO(n^{-\e}))\cdot d^2 \tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right], $$
	we can get from Lemma \ref{prop_model_shift_tight} that
\be\label{Deltabeta} 
\al_-^2(\rho_1) - \oo(1)  \le \frac{\delta_{\bias}}{ \Delta_{\bias}} \le \al_+^2(\rho_1) +  \oo(1) , \ee
	where 
	$$\Delta_{\bias}:=pd^2 \cdot \frac{\rho_1^2}{(\rho_1+\rho_2)^2}  \frac{1 + a_3 + a_4}{(a_1 + a_2)^2}= pd^2 \cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}.$$	
%\begin{align*}
%		\Delta_{\bias} &\define \hat{v}^2 \bignorm{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1 (\beta_s - \hat{v}\beta_t)}^2 \\
%		&= d^2 \cdot \bignormFro{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}X_1^{\top}X_1}^2 + \bigo{p^{-1/2 + \varepsilon}d^2}
%	\end{align*}
%	Using Lemma \ref{prop_model_shift_tight}, we get an upper and lower bound on $\Delta_{\bias}$ as
%	\be\label{Deltabeta} \bigbrace{1 - \sqrt{\frac{1}{c_1}}}^4 \le \Delta_{\bias} / \bigbrace{p\cdot d^2 \cdot \frac{c_1^2}{(c_1+c_2)^2} \cdot \frac{1 + a_3 + a_4}{(a_1 + a_2)^2} + \bigo{p^{1/2+\e}d^2} } \le \bigbrace{1 + \sqrt{\frac{1}{c_1}}}^4. \ee	
%	Hence we obtain that
%	\begin{align*}
%		\frac{1 + a_3 + a_4}{(a_1 + a_2)^2}
%		= \frac{(c_1 + c_2)^3}{(c_1 + c_2 - 1)^3} + \bigo{p^{-1/2+\varepsilon}}.
%	\end{align*}
Note that by \eqref{simple1}, we have
 \be\label{var-beta}\te(\hat{\beta}_2^{\STL})-\te(\hat{\beta}_2^{\MTL}) =\delta_{\vari} - \delta_{\bias} .\ee
Then we can track its sign using \eqref{Deltavar} and \eqref{Deltabeta}.

### 2020-09-27 arxiv/setup.tex
revise

\paragraph{Linear models.}
Suppose we have $t$ labeled regression tasks. %where $t$ is a fixed value that does not grow with the feature dimension $p$.
For each task $k$ from $1$ to $t$, we have $n_k$ samples $x_1, x_2, \dots, x_{n_k}$ that are all $p$-dimensional feature vectors with real-valued labels $y_1, y_2, \dots, y_{n_k}$.
Without loss of generality, we focus on predicting the $t$-th task and refer the $t$-th task as the target.
We refer to task $1$ to task $t-1$ as source tasks.
We assume that the labels of each task $k$ satisfy the linear model with unknown parameters $\beta_k\in\real^p$, that is,
	\[ y_i = x_i^{\top}\beta_k + e_i, \text{ for all } 1 \le i \le n_k, \]
where $e_i$ denotes random noise with mean zero and variance $\sigma^2$.
To write the above notations more succiently, let $X_k \in \real^{n_k \times p}$ denote the covariates that consists of a feature vector in every row.
Let
	\[ Y_k = X_k \beta_k + \varepsilon_k, \text{ for all } 1\le k \le t \] denote the label vector of task $k$ and $\varepsilon_k$ denote an i.i.d. random vector with mean zero and variance $\sigma^2$.

>>>>

\paragraph{Linear models.}
Suppose we have $t$ labeled regression tasks. %where $t$ is a fixed value that does not grow with the feature dimension $p$.
For each task $k$ from $1$ to $t$, we have $n_k$ samples $x_1, x_2, \dots, x_{n_k}$ that are all $p$-dimensional feature vectors with real-valued labels $y_1, y_2, \dots, y_{n_k}$.
Without loss of generality, we focus on predicting the $t$-th task and refer the $t$-th task as the target.
We refer to task $1$ to task $t-1$ as source tasks.
We assume that the labels of each task $k$ satisfy the linear model with unknown parameters $\beta_k\in\real^p$, that is,
	\[ y_i = x_i^{\top}\beta_k + e_i, \text{ for all } 1 \le i \le n_k, \]
where $e_i$ denotes random noise with mean zero and variance $\sigma^2$.
To write the above notations more succiently, let $X_k \in \real^{n_k \times p}$ denote the covariates that consists of a feature vector in every row.
Let
	\be\label{model_YvsX} Y^{(k)} = X^{(k)} \beta_k + \varepsilon_k, \text{ for all } 1\le k \le t \ee denote the label vector of task $k$ and $\varepsilon_k$ denote an i.i.d. random vector with mean zero and variance $\sigma^2$.
